---
title: "A ggplot primer"
draft: no
date: "2019-01-01"
lastmod: "2019-03-17"
status: evergreen
summary: Using one of my assignments to demonstrate ggplot usage
tags:
- psychology
- learning
---


## Brief description of the data 

The data within this document describes a repeated measures experiment whereby participants were allocated to four conditions. Initially, participants were briefly shown either a letter or a number. Then participants had to either respond to a target letter or number. The study measured each conditions' reaction time. Overall, the four conditions that each participant took part in were (alphabetical order):

Expect Letter -> Target Letter

Expect Letter -> Target Number

Expect Number -> Target Letter

Expect Number -> Target Number

## Setting up working environment

### Library loading
Here we use the library() function to load the packages needed for specific functions. For example, tidyverse has the package ggplot which allows for creating visualisations.

```{r, message = FALSE, error = FALSE, warning=FALSE}
library(tidyverse) # base library with tidyverse package
library(ggdist) # additional plots, e.g., half violin plots
library(ggforce) # additional plots, e.g., Sina plots
library(ggpubr) # publication ready themes
library(psych) # descriptive statistics package
library(patchwork)# allows for the combining of plots
library(knitr) # format tables
```

### Dataset importing
To conduct analysis, we need to import the dataset.
Using the read.csv function allows for the importing of a dataset. Here, the pathway is linked to an anonymous github account that viewers can access to examine the dataset.
Additionally, the concatenate function, "<-", assigns the imported data to the dataframe, "data".

We can inspect the data with functions such as head() or tail()
```{r}
data <- read.csv("https://raw.githubusercontent.com/ExcavatingMind/assignmentdata/main/datawranglingdataviz.csv") # read in csv from anonymous github repository

head(data) # view first 6 lines of dataset
```

## Data wrangling

### Renaming data

Currently, the names of the variables have no significance. Using the rename function allows for the conditions to be renamed into something more meaningful. For instance, ExpectN_LTarget means Expect number with letter target.

```{r}
renamed_data <- rename(data, ExpectN_NTarget = Condition1,
                       ExpectN_LTarget = Condition2, 
                       ExpectL_NTarget = Condition3,
                       ExpectL_LTarget = Condition4) # renaming each condition using the argument new name = old name
```

### Converting to long format

Most r and tidyverse functions work well with long data - each variable in a specific column. Here the pivot_longer() function allows for the condition columns to be combined into an IV: condition and DV: reaction time.

```{r}
long_data <- renamed_data %>%
  pivot_longer(cols = c(ExpectN_NTarget, ExpectN_LTarget, ExpectL_LTarget, 
                        ExpectL_NTarget),
               names_to = "Condition",
               values_to = "RT") # making two columns out of the original 4 with the IV being condiiton and the DV being reaction time
head(long_data)
```

### Separating the independent variable

Additionally, for some analyses, e.g., two way anova, it makes sense to split the Condition column into it's true levels (expect and target). Using the separate() function allows for a column to be separated based on a separator. In this case, it is the "_" symbol. However, when plotting graphs, the same visualisation can be produced for this data with either the long_data or the Interact_data as will be shown in the accessible academic visualisation section.

In addition, when expecting prior tibbles, we can see that the condition column is a character column. To conduct meaningful interaction analysis using ":", we need to mutate these variables into factors. Otherwise, ":" will not logically work. This is because the ":" symbol means from "x" to "y" whereby the x and y are factors (levelled constructs) or integers. For example 1:4 would mean 1,2,3,4.

```{r}
Interact_data <- long_data %>%
  separate(col = "Condition", into = c("Expect", "Target"), sep = "_") %>%  mutate(Expect = factor(Expect), Target = factor(Target)) # changing characters to factors; without this, the ":" operator will not work when examining interactions between two or more variables.

head(Interact_data)

```
## Summary statistics

### Manual summary statistics

Now we have our data in preferred formats, we need to generate summary statistics. One way to accomplish this is manually using the summarise() function. Within r, we can use formulas to produce outputs; for example, we can calculate the confidence interval value using a formula.

```{r , fig.dim = c(10,10)}
Summary_statistics <- long_data %>% 
  group_by(Condition) %>% # grouping the output by condition
  summarize(n = n(), # produce a count of participants
            mean = mean(RT), # produce mean of each condition
            sd = sd(RT), # produce standard deviation
            se = sd / sqrt(n), # produce standard error
            ci = qt(0.975, df = n - 1) * sd / sqrt(n), # confidence interval formula
            lci = mean - ci, # lower confidence interval
            uci = mean + ci) # upper confidence interval
kable(Summary_statistics) # display summary stats in a formatted table using knitr
```
### Using the psych package

While manually producing summary statistics is useful, it can be somewhat time-consuming. Using the psych package allows us to generate extensive summary statistics with less than three full lines of code.

```{r}
long_data2 <- long_data %>% # creating new tibble
   select(-ID) # removing ID as it is redundant for summary stats in the psych package
  
  describeBy(long_data2, long_data2$Condition) # creating summary statistics in psych package. The "$" symbol extracts a subset of the data. Here, we want to describe the data and separate the output by condition
```
## Data visualisation

### Basic distribution plots

#### Renaming factors

A common way to use data visualisation is to examine how normal the data is. However, when we are generating facetted visualisations, the variable names will be in r script format with "_" and no spaces. Here, we can rename the variables and create a new dataframe so that we do not have to rename each graph. 

```{r}
long_data_Q <- long_data # new tibble
  
  
  long_data_Q$Condition <- factor(long_data_Q$Condition,
                                levels = c("ExpectL_LTarget",
                                           "ExpectL_NTarget",
                                           "ExpectN_LTarget",
                                           "ExpectN_NTarget"),
                           labels = c("Expect Letter: Letter Target",
                                      "Expect Letter: Number Target",
                                      "Expect Number: Letter Target",
                                      "Expect Number: Number Target"))
# this argument renames the levels of the Condition variable with the argument levels = old name, labels = new name.
```

#### QQ-plot code

One useful plot is a qq-plot. A qq-plot can show how much skewness and kurtosis is in the data; essentially, it can demonstrate whether a dataset is normally distributed. Here we will use the ggplot package to construct a qq-plot.


```{r}
p1 <- long_data_Q %>%
  ggplot(aes(sample = RT, colour = Condition)) + # plot parameters
  stat_qq()  +  # qqplot
  stat_qq_line() + # qq line
  scale_color_brewer(palette = "Spectral") + # preselected colours
  theme_pubr() + # publication theme from ggpubr
  theme(text = element_text(family = "sans")) + # change font
  theme(panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank()) + # removing vertical gridlines
  theme(axis.text.x = element_text(size = 8)) + # axis font size
  guides(colour = "none") + #remove legend
  labs(title = "QQ-plot of each condition's reaction time scores", 
       y = "Reaction Time (ms)") + # labels for each component
facet_grid(~ Condition) # split graph by condition and put into grid format

```

#### Frequency polygon code

Another useful plot is the frequency polygon. It is similar to a histogram but visualises zero as a value so that we can see the specific counts and distribution of a dataset.


```{r}
p2 <- long_data_Q %>%
  ggplot(aes(x = RT)) +
  geom_freqpoly(aes(colour = Condition), 
                size = 1.5, # change size of lines
                binwidth = 10) +  
  theme_pubr() + 
  theme(text = element_text(family = "sans")) +
  scale_color_brewer(palette = "Set2") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  theme(axis.text.x = element_text(size = 8)) +
  guides(colour = "none") +
  labs(title = "Frequeny polygon of each condition's reaction time scores",
       x = "Reaction Time (ms)", 
       y = "Count") +
  facet_wrap(~ Condition) # make four graphs
```

#### Patchwork: combing two plots

Here, we have assigned the ggplot outputs to p1 and p2. We can use the package "patchwork" to weave two or more plots together. For instance, you could use the + operator to put them side by side or the / operator to put one above the other.

```{r , fig.dim= c(12, 12)}
p1 / p2 # patchwork package; using / places on graph above the other
```

#### Confirming visualisations with Shapiro-Wilks

Based on the psych package summary statistics and the visualisations, we can see that the condition "Expect Number Letter Target" displays moderate skew. To further examine if this is the case, i.e., if the condition is normally distributed, we could employ the shapiro wilk test.

```{r}
SW <- split(long_data2, long_data2$Condition) # making a vector of the dependent variable for each condition


EL_LT <- SW$ExpectL_LTarget$RT #splitting the data further by selecting only a specific condition's reaction time scores
EL_NT <-SW$ExpectL_NTarget$RT
EN_LT <-SW$ExpectN_LTarget$RT
EN_NT <-SW$ExpectN_NTarget$RT


shapiro.test(EL_LT) # shapiro-wilk test to examine normality
shapiro.test(EL_NT)
shapiro.test(EN_LT)
shapiro.test(EN_NT)
```
Indeed, the Shapiro-wilk test suggests that the Expect Number Letter Target condition violtes normality (***p*** = .028).

### Accessible academic visualisations

#### Violin-Box-Sina plot

It is important to make visualisations accessible to different populations and provide them with useful information. One way to do this in ggplot is to combine plots and build them onto eachother. A common plot combines the violin plot, boxplot, and jitter plot. However, the jitters display little information except count. Here, we can employ the ggforce package and use sina-plots. Sina-plots further demonstrate spread and density of points and provide more information. Overall, the violin-box-sina plot provides a range of summary statistics in visual form.

Additionally, this plot uses a colour scheme that is accessible to people with varying levels of colour blindness.

```{r , fig.dim = c(12, 6)}
Interact_data %>%
  ggplot(aes(x = Expect:Target, y = RT, fill = Expect:Target)) + 
  geom_violin() + # violin plot
  scale_fill_manual(breaks = c("ExpectL:LTarget", "ExpectL:NTarget", 
                               "ExpectN:LTarget", "ExpectN:NTarget"), 
                    values = c("#E69F00", 
                               "#F0E442", 
                               "#CC79A7", "#D55E00")) + # allows for the manual changing of colours
  geom_sina(alpha = .8, colour = "#0072B2") + # manually change colour of sina plots
  geom_boxplot(alpha = 4, # changes transparency
               outlier.shape = NA, # removes outliers 
               notch = T, # adds notches  
               colour = "black",
               width = .1, # changing boxplot width
               position = position_dodge(.9)) + # moving the position of the boxplot
  scale_x_discrete(labels = c("ExpectN:NTarget"= "Expect Number: Number Target", "ExpectN:LTarget" = "Expect Number: Letter Target", "ExpectL:LTarget" = "Expect Letter: Letter Target","ExpectL:NTarget" = "Expect Letter: Number Target")) + # renaming x axis labels
  guides(fill = "none") + 
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  theme(plot.title = element_text(hjust = 0.4), #  horizontally adjusting title
        plot.subtitle = element_text(hjust = .4)) + # horizontally adjusting subtitle
  theme(text = element_text(family = "sans")) +     
  theme(axis.text.x = element_text(size = 9)) +
  labs(title = "The effect of expectation and target on reaction time", 
       subtitle = "Receiving an expected target improves reaction time",
       x = "Condition", 
       y = "Reaction Time (ms)",
       caption = "Colour blind palette") # adds a caption
```

#### Raincloud plot

While the violin-box-sina plot provides useful information, it is relatively uncommon and many viewers may struggle to understand it. Another plot that is more accessible is the raincloud plot. Here, a half-violin plot is combined with box plots and dotplots. It produces the same information except it is slightly more viewer friendly. In essence, the half-violin plot acts as a variation of a histogram/frequency polygon which more people are familiar with.


```{r , fig.dim = c(12, 6)}

long_data %>% ggplot(aes(x = Condition, y = RT, fill = Condition)) + 
  ggdist::stat_halfeye(adjust = .5, # adjusts position 
                       width = 1, 
                       justification = -.1, # moves position left/right
                       .width = 0, # width of slab interval
                       point_colour = NA) + # colour of outlier points
  geom_boxplot(width = .12,
               outlier.color = "black") + # outlier colour
  ggdist::stat_dots(side = "left", # side of dots 
                    dotsize = 6, 
                    justification = 1.1, # adjusting left/right
                    binwidth = 1) +
  scale_fill_manual(breaks = c("ExpectN_NTarget", 
                               "ExpectL_NTarget", 
                               "ExpectL_LTarget", 
                               "ExpectN_LTarget"), 
                    values = c("#0066ff", 
                               "#FF66FF", 
                               "#33FF66", 
                               "#FF0000")) +
  scale_x_discrete(limits = c("ExpectL_LTarget", 
                              "ExpectN_LTarget", 
                              "ExpectN_NTarget", 
                              "ExpectL_NTarget"),
                   labels = c("ExpectN_NTarget" = "Expect Number with Number Target", "ExpectN_LTarget" = "Expect Number with Letter Target",
"ExpectL_LTarget" = "Expect Letter with Letter Target","ExpectL_NTarget" = "Expect Letter with Number Target")) +
  guides(fill = "none") + 
  theme_pubclean() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  theme(axis.text.x = element_text(size = 8)) +
  labs(title = "The effect of expectation and target on reaction Time", 
       x = "Condition", 
       y = "Reaction Time (ms)") +
  coord_flip()
```

#### Layperson lollipop plot

Finally, there are more people in the world that do not understand data visualisations than people who do. While it is important to provide useful information to laypeople, it must be balanced with appropriate systems, e.g., annotations. Here is an adapted lollipop plot with annotations and arrows to effectively communicate the data to laypeople.

The first stage is to create an absolute mean score that we can make into a horizontal line. This line would give reference to what the average score is across conditions.

```{r}

mean_avg <-
  long_data %>%
  summarise(avg = mean(RT)) %>%
  pull(avg) # creating an absolute mean score to add a line in a graph

```

Secondly, we need to add an additional column so our lollipop line ends at the lollipop and does not spill over.

```{r}
long_data_3 <- long_data %>%
  group_by(Condition) %>%
  mutate(mean_RT = mean(RT)) # adding mean scores for each condition as a column
```

Finally, we can construct the plot to produce an accessible data visualisation.

```{r , fig.dim = c(12, 10)}
long_data_3 %>%
  ggplot(aes(x = Condition, y = RT, colour = Condition)) +
  geom_hline(aes(yintercept = mean_avg), # where the horizontal absolute mean line parses through
             color = "gray70", # colour of line
             size = 0.6) + 
  stat_summary(fun = mean, geom = "point", size = 5) + # adding a large mean point for each condition; the lollipop 
  geom_sina(size = 2, alpha = .5) +
  geom_segment(aes(x = Condition, #coordinates for lollipop stick line 
                   xend = Condition,
                   y = mean_avg, 
                   yend = mean_RT), 
               size = .8) +
  annotate("text", 
           x = 3.7, # x,y co-ordinates for text annotations
           y = 1190, 
           size = 4, 
           color = "black", 
           lineheight = .9,
    label = "Mean reaction time of all conditions") + # the label for the annotation 
  annotate("text", x = 1, 
           y = 1290, 
           size = 4, 
           color = "gray20",
    label = "Mean of each condition") +
  annotate("text", 
           x = 2.7, 
           y = 1100, 
           size = 4, 
           color = "gray20",
    label = "Lowest mean reaction time of 995.42 ms") +
  annotate("text", 
           x = 1.9, 
           y = 910, 
           size = 4, 
           color = "gray20",
    label = "Sina plot shows distribution and density of scores") +
  geom_curve(aes(x = 1, # coordinate for arrows start and end positions 
                 y = 1280, 
                 xend = 1.95, 
                 yend = 1250), 
             arrow = arrow(length = unit(0.08, "inch")), # adding arrowhead
             size = 0.5, 
             colour = "black") +
  geom_curve(aes(x = 2.7,
                 y = 1090, 
                 xend = 3.95, 
                 yend = 990), 
             arrow = arrow(length = unit(0.08, "inch")),
             size = 0.5,
             colour = "black") +
  geom_segment(aes(x = 3.6,
                   y = 1180,
                   xend = 3.4,
                   yend = 1125), 
               arrow = arrow(length = unit(0.08, "inch")),
               size = 0.5,
               colour = "black") +
  geom_segment(aes(x = 1.8,
                   y = 920,
                   xend = 1.2,
                   yend = 970), 
               arrow = arrow(length = unit(0.08, "inch")),
               size = 0.5,
               colour = "black") +
  scale_x_discrete(labels = c("ExpectN_NTarget" = "Expect Number with Number Target", "ExpectN_LTarget" = "Expect Number with Letter Target",
"ExpectL_LTarget" = "Expect Letter with Letter Target",
"ExpectL_NTarget" = "Expect Letter with Number Target")) +
  scale_color_brewer(palette = "Set1") +
  guides(colour = "none") +
  theme_pubr() +
  theme(text = element_text(family = "sans")) +
  theme(title = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18)) + # change font size
  theme(axis.title.x = element_text(vjust = -.2)) + # move title down
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  theme(axis.text.x = element_text(size = 10)) +
  labs(title = "The effect of expectation and target on reaction time",
       subtitle = "Reaction time improves when expectation and target are the same",
       x = "Condition", 
       y = "Reaction Time (ms)")

```